\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-02-04 src/010chap1.tex
first draft of chapter 4 done. and hopefully the last draft.

\begin{definition}[Interval graph, Path graph]
  \label{def:pathgraph}
  Let $G$ be a graph.
  \begin{enumerate}
  \hangindent \defindent
  \item $G$ is an \textbf{interval graph} if there a set of intervals
    $\cI$ \stt $G$ is isomorphic to the intersection graph of
    $\cI$. \[G \cong \bI(\cI)\]
  \item $G$ is a \textbf{path graph} if there exists a tree $T$ and a
    set of paths from $T$, $\cP$ \stt $G$ is isomorphic to the
    intersection graph of $\cP$. \[G \cong \bI(\cP)\]
  \item $G$ is a \textbf{chordal graph} if it has no induced cycle of
    length greater than 3. Moreover, a \textbf{chordal graph} is
    characterized as an intersection graph of subtrees of a tree.
  \end{enumerate}
\end{definition}

>>>>

\begin{definition}[Interval graph, Path graph]
  \label{def:pathgraph}
  Let $G$ be a graph.
  \begin{enumerate}
  \hangindent \defindent
  \item $G$ is an \textbf{interval graph} if there a set of intervals
    $\cI$ \stt $G$ is isomorphic to the intersection graph of
    $\cI$. \[G \cong \bI(\cI)\]
  \item $G$ is a \textbf{path graph} if there exists a (undirected
    unrooted) tree $T$ and a set of paths from $T$, $\cP$ \stt $G$ is
    isomorphic to the intersection graph of $\cP$. \[G \cong
    \bI(\cP)\] Path graphs are also known as \uvgraphs (Undirected
    tree where paths intersect in a Vertex) \cite{mw86}.
  \item $G$ is a \textbf{chordal graph} if it has no induced cycle of
    length greater than 3. Moreover, a \textbf{chordal graph} is
    characterized as an intersection graph of subtrees of a tree.
  \end{enumerate}
\end{definition}

The set of interval graphs is a subset of set of path graphs which is
a subset of set of chordal graphs. In fact, there is another class
between chordal graphs and path graphs called RDV-graphs (Rooted
Directed tree where paths intersect in a Vertex)  \cite{mw86} which is
defined exactly as a path graph except that the trees are rooted and
directed.

### 2012-02-04 src/020chap2.tex
first draft of chapter 4 done. and hopefully the last draft.

\subsection{Finding forbidden submatrices}
It is known from Theorem~\ref{th:tuckercop}
that when a matrix does not have \COP, it must have one of Tucker's
forbidden submatrices. However, none of the \COP test algorithms discussed in
Section~\ref{sec:surveycoptest} outputs the forbidden submatrix when
the test is negative. This certainly gives a hint to how hard it is to find
these substructures in the input. \tnote[dire]{Fill neg results on
  finding forbidden submartrices -- boo75, some highlighted stuff from v85}

>>>>

\subsection{Finding forbidden submatrices}
\label{sec:forbiddensubalgo}

\tnote[dire]{
To study the smallest hamming distance \temptext{(?)} between the
input matrix and one with \COP, identifying the Tucker's forbidden
substructures (See Section~\ref{sec:tucker} for detailed
  discussion on Tucker's forbidden substructures for matrices with
  \COP) is of prime importance and we saw an algorithm for it in
  Section~\ref{sec:forbiddensubalgo}. 
}

It is known from Theorem~\ref{th:tuckercop}
that when a matrix does not have \COP, it must have one of Tucker's
forbidden submatrices. However, none of the \COP test algorithms discussed in
Section~\ref{sec:surveycoptest} outputs the forbidden submatrix when
the test is negative. This certainly gives a hint to how hard it is to find
these substructures in the input. \tnote[dire]{Fill neg results on
  finding forbidden submartrices -- boo75, some highlighted stuff from v85}

### 2012-02-04 src/030chap3.tex
first draft of chapter 4 done. and hopefully the last draft.

  \caption[\figtabsize Hypergraphs and set systems]{\figtabsize
    Hypergraphs and set systems. $\cF_1$ and $\cF_2$ are isomorphic
    hypergraphs. The subscript of each element is such that if $\phi$
    is the hypergraph isomorphism, $\phi(x_i)=y_i$. $\cI$ and $\cP$
    are an interval system and a path system respectively which are both
    isomorphic to $\cF_1$ (and by transitivity, isomorphic to $\cF_2$). Here too the
    subscript of elements indicate the isomorphism bijection. $\cF_3$
    is a hypergraph that is not isomorphic to any of the other
    hypergraphs -- moreover, it is one that cannot have a feasible
    path labeling to any tree since it does not have the Helly's
    property anymore due to the addition of hyperedge $\{x_1,
    x_4\}$. }
  \label{fig:hypergraphs}
\end{figure}

@@

\remove{It is an interesting fact that for a matrix with the COP, the
  intersection graph of the corresponding set system is an interval
  graph.  A similar connection to a subclass of chordal graphs, and
  this subclass contains interval graphs, exists for the
  generalization of COP.  In this case, the intersection graph of the
  corresponding set system must be a path graph. Chordal graphs are of
  great significance, extensively studied, and have several
  applications.  One of the well known and interesting properties of a
  chordal graphs is its connection with intersection graphs
  cite{mcg04}. For every chordal graph, there exists a tree and a
  family of subtrees of this tree such that the intersection graph of
  this family is isomorphic to the chordal graph
  cite{plr70,gav78,bp93}.  Certain format of these trees are called
  clique trees cite{apy92} of the graph which is a compact
  representation of the chordal graph. There has also been work done
  on the generalization of clique trees to clique hypergraphs
  cite{km02}.  If the chordal graph can be represented as the
  intersection graph of paths in a tree, then the graph is called path
  graph cite{mcg04} or path chordal graph.  Therefore, it is clear
  that if there is a bijection from $U$ to $V(T)$ such that the sets
  map to paths, then the intersection graph of the set system is
  indeed a path chordal graph.  This is, however, only a necessary
  condition and can be checked efficiently, as path graph recognition
  is polynomial solvable cite{gav78,aas93}.  Indeed, it is possible to
  construct a set system and tree, such that the intersection graph is
  a path chordal graph, but there is no bijection between $U$ and
  $V(T)$ such that the sets map to paths.  This connection indeed
  suggests that our problem is indeed as hard as path graph
  isomorphism.  Further path graph isomorphism is known be
  isomorphism-complete, see for example cite{kklv10}. } In the second
part of this paper, we decompose our search for a bijection between
$U$ and $V(T)$ into subproblems.  Each subproblem is on a set system
in which for each set, there is another set in the set system with
which the intersection is {\em strict}- there is a non-empty
intersection, but neither is contained in the other.  This is in the
spirit of results in \cite{wlh02,nsnrs09} where to test for COP in a
given matrix, the COP problem is solved on an equivalent set of prime
matrices.  \annote{Our decomposition localizes the challenge of path
  graph isomorphism to two problems.}{HOW?}

Finally, we show that Tree Path Assignment is isomorphism-complete.
We also point out Consecutive Ones Testing is in Logspace from two
different results in the literature \cite{kklv10, mcc04}. To the best
of our knowledge this observation has not been made earlier.

>>>>

  \caption[\figtabsize Hypergraphs and set systems]{\figtabsize
    Hypergraphs and set systems. $\cF_1$ and $\cF_2$ are isomorphic
    hypergraphs. The subscript of each element is such that if $\phi$
    is the hypergraph isomorphism, $\phi(x_i)=y_i$. $\cI$ and $\cP$
    are an interval system and a path system respectively which are both
    isomorphic to $\cF_1$ (and by transitivity, isomorphic to $\cF_2$). Here too the
    subscript of elements indicate the isomorphism bijection. $\cF_3$
    is a hypergraph that is not isomorphic to any of the other
    hypergraphs -- moreover, it is one that cannot have a feasible
    path labeling to any tree since it does not have the Helly's
    property \cite[Pr.~4.7]{mcg04} anymore due to the addition of hyperedge $\{x_1,
    x_4\}$. }
  \label{fig:hypergraphs}
\end{figure}

@@

In the second part of this paper, we decompose our search for a
bijection between $U$ and $V(T)$ into subproblems.  Each subproblem is
on a set system in which for each set, there is another set in the set
system with which the intersection is {\em strict}- there is a
non-empty intersection, but neither is contained in the other.  This
is in the spirit of results in \cite{wlh02,nsnrs09} where to test for
COP in a given matrix, the COP problem is solved on an equivalent set
of prime matrices.

% Finally, we show that Tree Path Assignment is isomorphism-complete.
We also point out Consecutive Ones Testing is in Logspace from two
different results in the literature \cite{kklv10, mcc04}. To the best
of our knowledge this observation has not been made earlier.

### 2012-02-04 src/040chap4.tex
first draft of chapter 4 done. and hopefully the last draft.

\chapter{Conclusion}
\label{ch:conclusion}

\begin{center} {\tt MOVED FROM INTRO SUBSECTION IN ch:myresearch
    CHAPTER begin}
\end{center}

% \annote{
It is an interesting fact that for a matrix with the COP, the
intersection graph of the corresponding set system is an interval
graph. % }{the following text could be in Further Study/Conclusion}
A similar connection to a subclass of chordal graphs and a superclass
of interval graphs exists for the generalization of COP.  In this
case, the intersection graph of the corresponding set system must be a
{\em path graph}. Chordal graphs are of great significance, are
extensively studied, and have several applications.  One of the well
known and interesting properties of a chordal graphs is its connection
with intersection graphs \cite{mcg04}. For every chordal graph, there
exists a tree and a family of subtrees of this tree such that the
intersection graph of this family is isomorphic to the chordal graph
\cite{plr70,gav78,bp93}.  These trees when in a certain format, are
called {\em clique trees} \cite{apy92} of the chordal graph. This is a
compact representation of the chordal graph. There has also been work
done on the generalization of clique trees to clique hypergraphs
\cite{km02}.  If the chordal graph can be represented as the
intersection graph of paths in a tree, then the graph is called path
graph \cite{mcg04}.  Therefore, it is clear that if there is a
bijection from $U$ to $V(T)$ such that for every set, the elements in
it map to vertices of a unique path in $T$, then the intersection
graph of the set system is indeed a path graph.  However, this is only
a necessary condition and can be checked efficiently because path
graph recognition is polynomial time
solvable\cite{gav78,aas93}. Indeed, it is possible to construct a set
system and tree, such that the intersection graph is a path graph, but
there is no bijection between $U$ and $V(T)$ such that the sets map to
paths. Path graph isomorphism is known be isomorphism-complete, see
for example \cite{kklv10}. An interesting area of research would be to
see what this result tells us about the complexity of the tree path
labeling problem (not covered in this paper).

\begin{center} {\tt MOVED FROM INTRO SUBSECTION IN ch:myresearch
    CHAPTER\\ end}
\end{center}

We give a characterization for feasible tree path labeling of path
hypergraphs. The proof for this is constructive and computes a
feasibility bijection mapping vertices of the hypergraph to vertices
of the given target tree. This thesis also discovered an exponential
algorithm that computes a feasible TPL to a given hypergraph if it is
a path hypergraph.

Our results have close connections to recognition of path graphs and
to path graph isomorphism.  Graphs which can be represented as the
intersection graph of paths in a tree are called {\em path
  graphs}\cite{mcg04}. Thus, a hypergraph $\cF$ can be interpreted as
paths in a tree, if and only if the intersection graph of $\cF$ is a
{\em path graph}. Path graphs are a subclass of chordal graphs since
chordal graphs are characterized as the intersection graphs of
subtrees of a tree\cite{mcg04}.  Path graphs are well studied in the
literature \cite{plr70}--\cite{mcg04}\tnote{Check if cite:bp93 is
  indeed for path graphs}.  Path graph recognition can be done in
polynomial time\cite{gav78,aas93}.  Clearly, this is a necessary
condition in terms of the intersection graph of the input hypergraph
$\cF$ in \FTPL. However, one can easily obtain a counterexample to
show the insufficiency of this condition\tnote{Give a figure
  showing insuff.}.  Path graph isomorphism is known be
isomorphism-complete\cite{kklv10}. Therefore, \annote{it is
  unlikely that we can solve the problem of finding feasible path
  labeling $\cl$ for a given $\cF$ and }{WHY?} tree $T$.  It is
definitely interesting to classify the kinds of trees and hypergraphs
for which feasible path labelings can be\tnote{ELABORATE.}  found
efficiently.  \annote{These results would form a natural
  generalization of COP testing and interval graph isomorphism,
  culminating in Graph Isomorphism }{REWORD} itself.  To this effect
we consider TPL on {\kstar} and give a polynomial time solution for
the same.

While we address the computation of TPL for a given
hypergraph% \footnote{The term {\em hypergraph} is used
% instead {\em set systems} are used interchangably in this
% section. Hypergraphs that have TPLs on some target tree are called
% {\em path hypergraphs}.}
from a given target tree in this research, however optimization
problems in TPL, akin to problems in Section~\ref{sec:optcop} for COP,
is outside the scope of this thesis. Whether TPL on general trees is
solvable in $P$ remains open. So do optimization opportunities in TPL
(possible extensions to optimization for COP in the
Section~\ref{sec:optcop}.\tnote{link to section}).

% \newpage %DUMMY
\tnote{ ||Complexity challenges:|| Canonization is an important
  tool in graph isomorphism. Invention of a
  % deterministic method of canonization for any class of graphs
  % naturally results in an algorithm for isomorphism. All that is
  % required is to check if the canons of two graphs are the
  % same. Thus complexity of graph isomorphism can be studied by
  % studying canonization methods. While general graph ismorphism
  % remains elusive in terms of hardness, canonization has been
  % studied for smaller classes of graphs thus giving
  % complexity/hardness results for them.

  % In 1992, \cite{sl92} made an important discovery that tree
  % isomorphism is in logspace. It was done by inventing a method of
  % canonization of trees using a logspace depth first traversal
  % algorithm. \tnote{confirm the statement of claim}.

  % \cite{kklv10} proved that interval graph canonization is also in
  % logspace thus drawing logspace conslusions about COT.
  % % This result is
  % % built on the fact that
  % Interval graphs are FP+C (fixed point with counting) definable
  % \tnote{elucidate} and \cite{l10} showed that this implies that it
  % captures PTIME. \tnote{elucidate} This result along with that of
  % undirected graph reachability being in logspace \cite{rei08},
  % \cite{kklv10} proved their logspace result.
}%
\tnote{ (1) the interesting survey in conclusion section of kklv10
  (2) [[[COP helping some problems with hardness]]] having COP in the
  input structure makes some problems less hard than in a general
  input. -- min set cover problem is known to be NP hard. so is min
  hitting
  set. these two problems are equivalent cite-adp80\\
  -- -- complexity results: log factor polynomial time approximable cite-f98, W[2]-complete with solution size as parameter cite-df99\\
  -- a couple of generalizations:\\
  -- -- min degree hypergraph (MDH) problem\\
  -- -- red blue covering (RBSC) problem\\
  ----- cite-d08phd Chapter 5.\\
  ----- what are cite-d08phd results\\
  ----- Rectangle stabbing \\
  ----- see cite-d08phd Chapter 6?  }

>>>>

\xclearpage

\def\p{{\sc P}\xspace}
\def\gicomplete{{\sc GI}-complete\xspace}
\def\logspace{{\sc logspace\xspace}\xspace}
\def\ptime{{\sc ptime}\xspace}
\def\fpplusc{{\sc FP+C}\xspace}

\chapter[Conclusion]{Conclusion and Further Research}
\label{ch:conclusion}

In this thesis, we surveyed \cop testing algorithms and as new results
we gave a characterization for a generalization of \COP, namely,
feasible tree path labeling of path hypergraphs. The proof for this is
constructive and computes a feasibility bijection mapping vertices of
the hypergraph to vertices of the given target tree. We also described
an exponential algorithm that computes a feasible TPL to a given
hypergraph if it is a path hypergraph.  In this chapter we will see a
few ideas where tree path labeling, in our experience, shows promising
scope for futher research.

\parindent 0pt

\textbf{Graph isomorphism and logspace canonization.}  Canonization is
an important tool in graph isomorphism. Invention of a deterministic
method of canonization for any class of graphs naturally results in an
algorithm for isomorphism. All that is required is to check if the
canons of two graphs are the same. Thus one way to study the
complexity of graph isomorphism is by studying canonization
methods. While general graph ismorphism remains elusive in terms of
hardness, canonization has been studied for smaller classes of graphs
thus giving complexity or hardness results for them -- some of the
canonization results are \cite{sl92,dlntw09,adkk09,kklv10}.

Ten years ago, \cite{sl92} made an important discovery that tree
isomorphism is in \logspace. It was done by inventing a method of
canonization of trees using a logspace depth first traversal
algorithm.  Less than a couple of years ago, \cite{kklv10} proved that
interval graph canonization is also in \logspace thus drawing logspace
conclusions about \COT. Interval graphs are \fpplusc (fixed point
logic with counting) definable and \cite{l10} showed that this implies
that it captures \ptime. This result along with that of undirected
graph reachability being in logspace \cite{rei08} helped \cite{kklv10}
conclude their logspace result.

How the work in this thesis connects to graph isomorphism and
canonization is as follows. For a matrix with the \COP, the
intersection graph of the input hypergraph is an interval graph.  A
similar connection exists for \TPL to path graphs which is a subclass
of chordal graphs and a superclass of interval graphs. Graphs which
can be represented as the intersection graph of paths in a tree are
path graphs (as seen in Section~\ref{sec:basicprelim}).  % In TPL, the
% intersection graph of the input hypergraph must be a path graph.  % Our
% results have close connections to recognition of path graphs and to
% path graph isomorphism.  
Thus, a hypergraph $\cF$ can be labeled with paths in a tree (path
labeling as defined in Section~\ref{ch:prelims}), \iff the
intersection graph $\bI(\cF)$ is a path graph. Path graphs are a
subclass of chordal graphs since chordal graphs are characterized as
the intersection graphs of subtrees of a tree \cite[Ch.~4,
Sec.~5]{mcg04}. Chordal graphs are of great significance, are
extensively studied, and have several applications. For every chordal
graph, there exists a tree and a family of subtrees of this tree such
that the intersection graph of this family is isomorphic to the
chordal graph \cite{jrw72phd,gav74b,bun74}.  These trees when in a
certain format, are called {\em clique trees} \cite{ppy92} of the
chordal graph. This is a compact representation of the chordal
graph. There has also been work done on the generalization of clique
trees to clique hypergraphs \cite{km02}.  Path graphs are well studied
in the literature (characterization for \uvgraphs was perhaps first
mentioned by \cite{plr70}).
% Clearly, this is a necessary condition in terms of the intersection
% graph of the input hypergraph $\cF$ in \FTPL. However, one can easily
% obtain a counterexample to show the insufficiency of this
% condition\tnote{Give a figure showing insuff.}.
Indeed, path graph recognition is a necessary condition for \CFTPL but
not sufficient.  We know that path graph recognition can be done in
polynomial time \cite[first algorithm.
$O\left(pn^3\right)$]{gav78},\cite[$O\left(p(m+n)\right)$]{aas93}
\cite{sb94}, \cite[linear algorithm]{db95}, \cite[PR-trees]{cha11}
\footnote{$n$ is number of vertices, $m$ is number of edges, $p$ is
  number of maximal cliques in the input graph.}.  Path graph
isomorphism is also known be \gicomplete \cite{bpt96}.  It would be
interesting to pursue research to answer the following questions in
relation to TPL and graph isomorphism.
\begin{enumerate}
\item What does path graph isomorphism hardness tell us about the
  complexity of \CFTPL?
\item Can TPL help in computing a canon for path graphs, especially a
  logspace canon?
\end{enumerate}

%\noindent
\textbf{Complexity of TPL in other tree classes.}  We know that on
intervals, \CFTPL is easy (Section~\ref{sec:icpplicpia}).  We showed
that \CFTPL on {\kstar} is also polynomial time solvable in
Section~\ref{sec:ksubdivstar}. However, due to its close connections
with path graph isomorphism which is \gicomplete \cite{kklv10}, we
conjecture that \CFTPL on general trees cannot do better than
\gicomplete. Our algorithm in Section~\ref{sec:decompos} is unable to
solve it in \p. It would therefore, be certainly of interest to
classify the kinds of trees and hypergraphs for which feasible path
labelings can be found efficiently, or better still, to solve \CFTPL
for general trees itself in \p-time or prove that it is not in \p.

%\noindent
\textbf{Optimization problems in TPL.}  There are optimization
problems for matrices that do not have \COP which we only mention in
Section~\ref{sec:optcop} but did not discuss further since it is not
in the scope of the main work of this thesis. We mention them here
again to encourage further analogous research in optimization problems
of \TPL.  The central questions in this area are (i) how close is the
matrix to having \COP, mainly in terms of Tucker's forbidden
submatrices, and (ii) how optimally can one alter the matrix to attain
\COP.  With regard to the latter question, recent literature indicate
that there has been a lot of interest in matrix modification problems
to make a matrix have \COP \cite{hg02, tz07}. While we address the
computation of TPL for a given hypergraph from a given target tree in
this research, however optimization opportunities in TPL are yet to be
explored and their complexity remains open. We suggest a few
optimization ideas here and there could be many more such questions
that could be asked by a motivated researcher. 
\begin{enumerate}
\item If \CFTPL is not possible with the input target tree $T$, 
  can $T$ be changed to get a feasible TPL?
\item What is the characterization of a hypergraph that will fail the
  \CFTPL problem for any target tree? \ie what is the characterization
  of a hypergraph that is not a path hypergraph? Can a forbidden
  substructure be found?
\item If a given hypergraph is not a path hypergraph, can at most $k$
  vertices be deleted from the hypergraph to make it a path
  hypergraph? A dual problem would be to select at least $k$ vertices
  to induce a path hypergraph? A different flavour of the problem
  would be deleting (or selecting) hyperedges in the input to get a
  path hypergraph (analogous to {\sc Min COS-R, Min COS-C, Max
    COS-R, Max COS-C} problems in \COP).
\item Can at most $k$ vertex memberships in hyperedges be
  reversed to get a path hypergraph? (analogous to {\sc COP
    Augmentation} problem).
\end{enumerate}
It is unlikely that these problems will have efficient exact
algorithms since their \COP counterparts are all NP hard
problems. Thus the scope of research would be to find approximation
algorithms or fixed parameter algorithms or hardness results.

% %\noindent

% \temptext{ 
%   (2) [[[COP helping some problems with hardness]]] having COP in the
%   input structure makes some problems less hard than in a general
%   input. -- min set cover problem is known to be NP hard. so is min
%   hitting
%   set. these two problems are equivalent cite-adp80\\
%   -- -- complexity results: log factor polynomial time approximable
%   cite-f98, W[2]-complete with solution size as parameter cite-df99\\ 
%   -- a couple of generalizations:\\
%   -- -- min degree hypergraph (MDH) problem\\
%   -- -- red blue covering (RBSC) problem\\
%   ----- cite-d08phd Chapter 5.\\
%   ----- what are cite-d08phd results\\
%   ----- Rectangle stabbing \\
%   ----- see cite-d08phd Chapter 6?  }

%\noindent
\textbf{x-dimensional generalization of \COP.} As seen in the chapters
of this thesis \TPL is a graph theoretic generalization of
\COP. Intervals are target trees with maximum degree 2 and general
trees have no bound on maximum degree. Another way to look at
intervals is using geometry. Intervals can be imagined in one
dimension as line segments along a single line in the Euclidean space
and the elements of hypergraph universe are coordinate points on this
line segment. The generalization of this visualization is obvious --
how does the interval labeling property generalize when each hyperedge
must be mapped to coordinate points bounded by a rectilinear 2-D shape
on a single plane, say rectangles (analogous to coordinate points on
line segments in interval labeling problem). One hardness result is
already known -- the intersection graphs of axis-parallel boxes in
$\bR^d$ is GI-complete \cite{ueh08}.
